---
title: "Anticipatory Framework for Pakistan Floods (2022)"
author: "Hackathon Team"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load required libraries
if (!require("pacman")) install.packages("pacman")
pacman::p_load(sf, terra, dplyr, readr, ggplot2, leaflet, exactextractr, data.table, lubridate, tidyr, purrr, patchwork, FactoMineR, factoextra)
```

## 1. Load Spatial Boundaries

We use the OCHA subnational boundaries for Pakistan at the ADM2 (District) level.

```{r load_spatial}
# Define base paths
data_dir <- "/Users/martinapardy/Desktop/hackathon_data/unzipped"

# Load ADM2 boundaries
adm2_path <- file.path(data_dir, "spatial_boundaries/spatial_boundaries/ocha_subnational_boundaries/pak_admbnda_adm2_wfp_20220909.shp")

# Read using sf
pak_adm2 <- st_read(adm2_path, quiet = TRUE)

# Ensure CRS is correct (WGS84)
pak_adm2 <- st_transform(pak_adm2, crs = 4326)

# Quick look at the boundaries
ggplot(data = pak_adm2) +
  geom_sf(fill = "lightblue", color = "white", size = 0.2) +
  theme_minimal() +
  labs(title = "Pakistan ADM2 Boundaries")
```

## 2. Flood Extent Mapping (2022)

Here we map the flood extent from the provided shapefiles for the 2022 floods to visualize the spatial distribution of the affected areas over the whole country.

```{r load_flood_extents, fig.width=10, fig.height=6}
# Define paths for 2022 
path_2022_flood <- file.path(data_dir, "flood_data/flood_data/FL20220808PAK_SHP/VIIRS_20220803_20220823_FloodExtent_PAK.shp")
path_2022_analysis <- file.path(data_dir, "flood_data/flood_data/FL20220808PAK_SHP/VIIRS_20220803_20220823_AnalysisExtent_PAK.shp")

# Read and transform the shapefiles
flood_2022 <- st_read(path_2022_flood, quiet = TRUE) %>% st_transform(crs = 4326)
analysis_2022 <- st_read(path_2022_analysis, quiet = TRUE) %>% st_transform(crs = 4326)

# Map 2022 Flood and Analysis Extent
plot_2022 <- ggplot() +
  geom_sf(data = pak_adm2, fill = "gray90", color = "white", size = 0.2) +
  geom_sf(data = analysis_2022, fill = "lightyellow", color = NA, alpha = 0.5) +
  geom_sf(data = flood_2022, fill = "blue", color = NA, alpha = 0.6) +
  theme_minimal() +
  labs(title = "Pakistan Flood Extent (2022)",
       subtitle = "VIIRS (03-23 Aug 2022)\nYellow: Analysis Extent")

plot_2022
```

## 3. Load Deprivation Data (Vulnerability)

We load the GRDI data which gives a baseline vulnerability score. The raw data is a grid (GeoTIFF raster format) with a spatial resolution of roughly 1 km. 

```{r load_deprivation}
# Load raw GRDI Raster (Grid Format)
grdi_raster_path <- file.path(data_dir, "deprivation_data/deprivation_data/raster/grdi_pakistan.tif")
grdi_raster <- rast(grdi_raster_path)

# Quick plot of the raster grid with most deprived areas in red
red_palette <- colorRampPalette(c("lightyellow", "orange", "darkred"))(100)
plot(grdi_raster, main = "Global Risk Data Index (GRDI) - Pakistan (1km Grid)", col = red_palette)

# Load GRDI ADM2 pre-aggregated tabular data
grdi_path <- file.path(data_dir, "deprivation_data/deprivation_data/aggregated/grdi_ocha_adm2.csv")
grdi_adm2 <- read_csv(grdi_path, show_col_types = FALSE)

# Join with spatial boundaries
pak_adm2_grdi <- pak_adm2 %>%
  left_join(grdi_adm2 %>% select(ADM2_PCODE, mean_rdi), by = "ADM2_PCODE")

# Plot the aggregated ADM2 version to show how it cleans up the noise
ggplot(data = pak_adm2_grdi) +
  geom_sf(aes(fill = mean_rdi), color = "gray20", size = 0.1) +
  scale_fill_gradientn(colors = c("lightyellow", "orange", "darkred"), na.value = "gray90", name = "Mean RDI") +
  theme_minimal() +
  labs(title = "Aggregated GRDI - Pakistan (District Level)",
       subtitle = "Averaged vulnerability score per ADM2 district")
```

## 4. Population Movement (Extracting an Anomaly)

We process the Facebook population data to extract anomaly signals (`z_score`).

**How do we detect an anomaly?**
We use **The "Statistically Significant" Threshold (Z < -2.0)**.
* **What it means:** The population has dropped more than 2 standard deviations below normal.
* **Is it an anomaly?** Yes. In statistics, a Z-score outside of ±2.0 means there is less than a 2.5% chance that this drop happened naturally. When a district hits a -2.0, humanitarian organizations start paying attention because the data is signaling that something highly unusual is forcing people to leave.

```{r process_mobility}
agg_dir <- file.path(data_dir, "facebook_population_movement_data/facebook_population_movement_data/facebook_data_floods_2022/population/aggregated")

# Get list of all csv files
csv_files <- list.files(agg_dir, pattern = "\\.csv$", full.names = TRUE)

# Read all files and combine into one dataframe
df <- map_df(csv_files, read_csv, show_col_types = FALSE)

# Clean up date and ensure numeric columns
df <- df %>%
  mutate(
    clipped_z_score = as.numeric(clipped_z_score),
    percent_change = as.numeric(percent_change),
    date_time = as.POSIXct(date_time, format="%Y-%m-%d %H:%M", tz="UTC"),
    date = as.Date(date_time),
    is_anomaly = ifelse(clipped_z_score < -2.0, 1, 0)
  )
```

### National Signal (Daily & Granular)

We aggregate the signal across the country to see macro-level displacement trends. We generate two views: a smoothed Daily Average, and a Granular view showing all 3 data collection timestamps per day.

```{r plot_national, fig.width=12, fig.height=8}
# 1. Daily Average
daily_national <- df %>%
  group_by(date) %>%
  summarize(
    mean_z_score = mean(clipped_z_score, na.rm = TRUE),
    mean_percent_change = mean(percent_change, na.rm = TRUE)
  )

plot_daily <- ggplot(daily_national, aes(x = date, y = mean_z_score)) +
  geom_line(linewidth = 1, color = "darkblue") +
  geom_point(size = 2, color = "darkblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  theme_minimal() +
  labs(
    title = "National Average (Daily)",
    subtitle = "Aggregated daily mean Z-Score",
    x = "Date",
    y = "Mean Clipped Z-Score"
  )

# 2. Granular (3 timestamps per day)
national_trend <- df %>%
  group_by(date_time) %>%
  summarize(
    mean_z_score = mean(clipped_z_score, na.rm = TRUE),
    mean_percent_change = mean(percent_change, na.rm = TRUE)
  )

plot_granular <- ggplot(national_trend, aes(x = date_time, y = mean_z_score)) +
  geom_line(linewidth = 1, color = "darkred") +
  geom_point(size = 2, color = "darkred") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  theme_minimal() +
  labs(
    title = "National Average (Granular)",
    subtitle = "3 timestamps per day across all districts",
    x = "Date & Time",
    y = "Mean Clipped Z-Score"
  )

plot_daily / plot_granular
```

### 5 Most Affected Districts (Granular Timestamps)

To better understand the localized severity, we isolate the 5 districts that experienced the most extreme population drops and plot their specific granular trends.

```{r plot_top5, fig.width=10, fig.height=6}
# 1. Identify the top 5 most affected districts based on the lowest Z-score recorded
top5_districts <- df %>%
  group_by(polygon_name) %>%
  summarize(min_z = min(clipped_z_score, na.rm = TRUE)) %>%
  arrange(min_z) %>%
  slice(1:5) %>%
  pull(polygon_name)

# 2. Filter the raw data for just those 5 districts using granular date_time
top5_data <- df %>%
  filter(polygon_name %in% top5_districts) %>%
  group_by(date_time, polygon_name) %>%
  summarize(mean_z_score = mean(clipped_z_score, na.rm = TRUE), .groups="drop")

# 3. Plot the individual district trends
ggplot(top5_data, aes(x = date_time, y = mean_z_score, color = polygon_name)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0, linetype = "solid", color = "black") +
  geom_hline(yintercept = -2, linetype = "dashed", color = "red") + # Anomaly threshold
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Population Anomaly Trends for Top 5 Affected Districts",
    subtitle = "Granular view (3 timestamps per day). Red dashed line = Extreme Anomaly Threshold.",
    x = "Date & Time",
    y = "Mean Clipped Z-Score",
    color = "District"
  )
```

## 5. Process Rainfall Data (Hazard Trigger)

Rainfall acts as the physical hazard trigger. We extract the total accumulated rainfall from the IMERG satellite data for each district.

```{r process_rainfall}
# Get list of daily rainfall TIFF files
rainfall_files <- list.files(
  path = file.path(data_dir, "rainfall_data/rainfall_data/rainfall_data_2022"), 
  pattern = "\\.tif$", 
  full.names = TRUE
)

# Load as a raster stack
rainfall_stack <- rast(rainfall_files)

# Extract mean daily rainfall per district
# This returns a dataframe where each column is a day's mean rainfall
district_rainfall <- exact_extract(rainfall_stack, pak_adm2_grdi, 'mean', progress = FALSE)

# Sum all days to get total accumulated rainfall per district
# We calculate this on the plain dataframe BEFORE binding it to avoid sf geometry errors
district_rainfall <- district_rainfall %>%
  mutate(total_rainfall = rowSums(select(., starts_with("mean")), na.rm = TRUE))

pak_adm2_grdi <- bind_cols(pak_adm2_grdi, district_rainfall)
```

## 6. Process NDVI Data (Antecedent Conditions)

Vegetation and soil moisture conditions prior to the hazard act as an amplifier for flood risk. If the soil is already saturated or vegetation is highly anomalous, floods are more severe. We load the historical NDVI data and extract the pre-flood anomaly index for each district.

```{r process_ndvi}
ndvi_path <- file.path(data_dir, "climate_data/climate_data/pak-ndvi-subnat-5ytd.csv")
ndvi_data <- read_csv(ndvi_path, show_col_types = FALSE)

# Filter for ADM2 (Districts) and extract the latest pre-flood anomaly
# Using viq (NDVI anomaly quotient) as the antecedent condition indicator
ndvi_district <- ndvi_data %>%
  filter(adm_level == 2) %>%
  mutate(date = as.Date(date)) %>%
  filter(date < as.Date("2025-08-01")) %>% # Pre-flood baseline
  group_by(PCODE) %>%
  arrange(desc(date)) %>%
  slice(1) %>% # Take the most recent reading before the flood
  ungroup() %>%
  select(ADM2_PCODE = PCODE, ndvi_anomaly = viq)

# Join to our master spatial dataframe
pak_adm2_grdi <- pak_adm2_grdi %>%
  left_join(ndvi_district, by = "ADM2_PCODE") %>%
  mutate(
    # Fill missing values with 100 (normal/no anomaly) to prevent PCA failure
    ndvi_anomaly = ifelse(is.na(ndvi_anomaly), 100, ndvi_anomaly)
  )
```

## 7. Deprivation, Mobility, and Flood Extent Visualization (2022)

Here we visualize the baseline deprivation (vulnerability) map alongside the high-resolution tile-level mobility signals and the actual flood water extent.

```{r dep_vs_mob_tiles_2022, fig.width=18, fig.height=7}
# Map 1: Deprivation (High Deprivation in Red, using Quintiles)
pak_adm2_grdi <- pak_adm2_grdi %>%
  mutate(
    deprivation_quintile = cut(
      mean_rdi, 
      breaks = quantile(mean_rdi, probs = seq(0, 1, by = 0.2), na.rm = TRUE),
      labels = c("Very Low", "Low", "Medium", "High", "Very High"),
      include.lowest = TRUE
    )
  )

map_deprivation <- ggplot(data = pak_adm2_grdi) +
  geom_sf(aes(fill = deprivation_quintile), color = "white", size = 0.1) +
  scale_fill_brewer(palette = "YlOrRd", name = "GRDI\n(Deprivation)", na.value = "gray90") +
  theme_minimal() +
  labs(title = "Baseline Deprivation (GRDI)",
       subtitle = "Categorized into Quintiles (Very Low to Very High)")

# --- 2022 Maps ---
fb_tiles_dir_2022 <- file.path(data_dir, "facebook_population_movement_data/facebook_population_movement_data/facebook_data_floods_2022/population/tile")
fb_files_2022 <- list.files(fb_tiles_dir_2022, pattern = "\\.csv$", full.names = TRUE)

if(length(fb_files_2022) > 0) {
  fb_data_2022 <- rbindlist(lapply(fb_files_2022[1:5], fread))
  fb_sf_2022 <- st_as_sf(fb_data_2022, coords = c("lon", "lat"), crs = 4326)
  
  anomalous_tiles_2022 <- fb_sf_2022 %>% filter(percent_change < 0)
  
  map_tiles_2022 <- ggplot() +
    geom_sf(data = pak_adm2, fill = "gray90", color = "white", size = 0.1) +
    geom_sf(data = anomalous_tiles_2022, aes(color = percent_change), size = 0.5, alpha = 0.7) +
    scale_color_viridis_c(option = "mako", direction = -1, name = "% Drop") +
    theme_minimal() +
    labs(title = "2022 Population Displacement (Tiles)",
         subtitle = "Tile-level population drops")
  
  map_deprivation + map_tiles_2022 + plot_2022 + plot_layout(ncol = 3)
}
```

## 8. Simple Anticipatory Risk Model (Manual Weighting)

Before trying a complex statistical model, we can build a simple index by manually weighting the variables. Here we combine Deprivation, Rainfall, and the active crisis trigger (Mobility Anomaly) with equal weights.

```{r simple_risk_model}
# Standardize deprivation, rainfall, and NDVI (0 to 1)
pak_adm2_grdi <- pak_adm2_grdi %>%
  mutate(
    vuln_idx = (mean_rdi - min(mean_rdi, na.rm=T)) / (max(mean_rdi, na.rm=T) - min(mean_rdi, na.rm=T)),
    rain_idx = (total_rainfall - min(total_rainfall, na.rm=T)) / (max(total_rainfall, na.rm=T) - min(total_rainfall, na.rm=T)),
    ndvi_idx = (ndvi_anomaly - min(ndvi_anomaly, na.rm=T)) / (max(ndvi_anomaly, na.rm=T) - min(ndvi_anomaly, na.rm=T))
  )

# Get latest mobility signal and check if there's an active anomaly
latest_date <- max(df$date, na.rm = TRUE)
latest_mobility <- df %>% 
  filter(date == latest_date) %>%
  group_by(polygon_name) %>%
  summarize(
    mean_z_score = mean(clipped_z_score, na.rm = TRUE),
    is_anomaly = max(is_anomaly, na.rm = TRUE) # 1 if anomalous, 0 if not
  )

simple_risk_df <- pak_adm2_grdi %>%
  left_join(latest_mobility, by = c("ADM2_EN" = "polygon_name")) %>%
  mutate(
    is_anomaly = ifelse(is.na(is_anomaly), 0, is_anomaly),
    
    # Calculate Risk Score (0-100) combining Vulnerability, Hazard (Rain), Antecedent (NDVI), and Trigger
    Simple_Risk_Score = (0.25 * vuln_idx + 0.25 * rain_idx + 0.25 * ndvi_idx + 0.25 * is_anomaly) * 100
  )

ggplot(data = simple_risk_df) +
  geom_sf(aes(fill = Simple_Risk_Score), color = "white", size = 0.1) +
  scale_fill_viridis_c(option = "magma", direction = -1, name = "Simple Risk\nScore") +
  theme_minimal() +
  labs(title = paste("Simple Composite Risk Score on", latest_date),
       subtitle = "Combines Deprivation (25%), Rainfall (25%), NDVI (25%), and Active Anomaly (25%)")
```

## 9. Advanced Anticipatory Risk Modeling (PCA)

We upgrade our manual weighting approach to an unsupervised Principal Component Analysis (PCA). By feeding Deprivation, Rainfall, and NDVI Anomaly into the PCA, the algorithm mathematically derives an optimized "Structural Risk Index" (PC1) based on the variance in the data. We then combine this structural index with our active Facebook Mobility trigger.

```{r pca_model}
# 1. Prepare data for PCA
pca_data <- pak_adm2_grdi %>%
  st_drop_geometry() %>%
  select(ADM2_PCODE, ADM2_EN, mean_rdi, total_rainfall, ndvi_anomaly) %>%
  na.omit() # PCA requires complete cases

# 2. Run PCA on the 3 structural variables
# Scale = TRUE is critical because these variables have completely different units
res.pca <- PCA(pca_data[, c("mean_rdi", "total_rainfall", "ndvi_anomaly")], scale.unit = TRUE, graph = FALSE)

# 3. Extract Principal Component 1 (The new Structural Risk Index)
# PC1 typically captures the most variance (overall severity)
pca_data$pca_risk_index <- res.pca$ind$coord[, 1]

# Standardize the PCA Index from 0 to 1 so it's easy to interpret
pca_data <- pca_data %>%
  mutate(
    pca_risk_index = (pca_risk_index - min(pca_risk_index)) / (max(pca_risk_index) - min(pca_risk_index))
  )

# 4. Join back to the spatial dataframe
pak_adm2_grdi <- pak_adm2_grdi %>%
  left_join(pca_data %>% select(ADM2_PCODE, pca_risk_index), by = "ADM2_PCODE")

# 5. Bring in the Facebook Anomaly Trigger
latest_date <- max(df$date, na.rm = TRUE)
latest_mobility <- df %>% 
  filter(date == latest_date) %>%
  group_by(polygon_name) %>%
  summarize(is_anomaly = max(is_anomaly, na.rm = TRUE))

risk_df <- pak_adm2_grdi %>%
  left_join(latest_mobility, by = c("ADM2_EN" = "polygon_name")) %>%
  mutate(
    is_anomaly = ifelse(is.na(is_anomaly), 0, is_anomaly),
    
    # Final Anticipatory Score: Combine the PCA structural risk with the Active Evacuation Trigger
    Final_Risk_Score = (0.5 * pca_risk_index + 0.5 * is_anomaly) * 100
  )

# Plot the Final PCA-driven Risk Score
ggplot(data = risk_df) +
  geom_sf(aes(fill = Final_Risk_Score), color = "white", size = 0.1) +
  scale_fill_viridis_c(option = "magma", direction = -1, name = "PCA Risk\nScore") +
  theme_minimal() +
  labs(title = paste("Advanced PCA Composite Risk Score on", latest_date),
       subtitle = "Combines PCA Structural Index (Deprivation + Rain + NDVI) with Active Evacuation Trigger")
```
